---
title: "NC/FET Estrodiol Statistic in R"
author: "Matthew Sahagun"
date: "5/25/2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, include = FALSE}
library(pacman)
p_load(gtools, knitr)
```

## Introduction

For this project, we want to know if estradiol values affect the prgnancy rates of women. To help with this study, below I will clean the data, run some single variable analysis, and then run logistic analysis. 

## Examining the Data

```{r}
estrodiol = read.csv("/Users/mjs13/Downloads/NC-FET Estradiol ASRM 5-2020 BB (6).csv", header = TRUE)
```
```{r}
head(estrodiol)
names(estrodiol)
```

## Cleaning the Data

I found this nifty outlier function that seems like it will be useful. 

```{r}
#outlier function
outlierKD <- function(dt, var) {
     var_name <- eval(substitute(var),eval(dt))
     na1 <- sum(is.na(var_name))
     m1 <- mean(var_name, na.rm = T)
     par(mfrow=c(2, 2), oma=c(0,0,3,0))
     boxplot(var_name, main="With outliers")
     hist(var_name, main="With outliers", xlab=NA, ylab=NA)
     outlier <- boxplot.stats(var_name)$out
     mo <- mean(outlier)
     var_name <- ifelse(var_name %in% outlier, NA, var_name)
     boxplot(var_name, main="Without outliers")
     hist(var_name, main="Without outliers", xlab=NA, ylab=NA)
     title("Outlier Check", outer=TRUE)
     na2 <- sum(is.na(var_name))
     cat("Outliers identified:", na2 - na1, "n")
     cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "n")
     cat("Mean of the outliers:", round(mo, 2), "n")
     m2 <- mean(var_name, na.rm = T)
     cat("Mean without removing outliers:", round(m1, 2), "n")
     cat("Mean if we remove outliers:", round(m2, 2), "n")
     response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
     if(response == "y" | response == "yes"){
          dt[as.character(substitute(var))] <- invisible(var_name)
          assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
          cat("Outliers successfully removed", "n")
          return(invisible(dt))
     } else{
          cat("Nothing changed", "n")
          return(invisible(var_name))
     }
}
```


I determined that all of the values in the column Estradiol.value are factors. I need them to be numeric. 

```{r}
#change columns from factor to numeric
estrodiol$Estradiol.value = as.numeric(as.character(estrodiol$Estradiol.value))
class(estrodiol$Estradiol.value)

estrodiol$BMI = as.numeric(as.character(estrodiol$BMI))
class(estrodiol$BMI)

estrodiol$Age.at.transfer = as.numeric(as.character(estrodiol$Age.at.transfer))
class(estrodiol$Age.at.transfer)
```

I am checking for missing value. In an earlier version of the data set, there were quite a few instances of missing values. That is no longer the case. I also wanted to look at the boxplot of the data, and noted many outliers. 
```{r}
#counting the number of NA values
sum(is.na(estrodiol$Estradiol.value))

#removing NA values since they are not relevant to our study. 
row = which(is.na(estrodiol$Estradiol.value))
row
#estrodiol = estrodiol[-row,]
#sum(is.na(estrodiol$Estradiol.value))

#boxplot to look at data
boxplot(estrodiol$Estradiol.value)

#Impute NA from Estrodiol.value column with median
#estrodiol$Estradiol.value[is.na(estrodiol$Estradiol.value)] = median(estrodiol$Estradiol.value, na.rm = T)

#finding mean of column
mean(estrodiol$Estradiol.value)

#outliers. We have 26 of them
outlier_values = boxplot.stats(estrodiol$Estradiol.value)$out
length(outlier_values)
sort(outlier_values)
#outlier_rows = 

outlierKD(estrodiol, Estradiol.value)
```

I went in search for the rows of the outliers. 

```{r}
#finding the rows of the outliers
row = array()
values = array()

for (i in 1:length(outlier_values)) {
  values[i] = outlier_values[i]
  row[i] = which(estrodiol$Estradiol.value == outlier_values[i])
}

dat = data.frame(row, values)
sorted_dat = dat[order(dat$values),]
sorted_dat
```

I was told by the researchers that those with an estradiol value greater than 708 were given supplements, which is not part of the study. I vill remove those values, which takes care of my outlier problem. 

```{r}
#removing rows where the estrodiol value is greater than 708, since those patients were given supplements. 
supp = which(estrodiol$Estradiol.value > 708)
supp
length(supp)

estrodiol = estrodiol[-supp,]
nrow(estrodiol)

#Our data now has 441 rows after removing the outliers (greater than 708 estrodiol level)
```

I will also categorize the ages in the following manner, and then add a column for these values into the dataframe: <30, 30-34, 35-39, >=40
```{r}
age_transfer_cat = array()

for (i in 1:nrow(estrodiol)) {
  if (estrodiol$Age.at.transfer[i] < 30) {
    age_transfer_cat[i] = "<30"
  } else if (estrodiol$Age.at.transfer[i] >= 30 & estrodiol$Age.at.transfer[i] < 35) {
    age_transfer_cat[i] = "30-34"
  } else if (estrodiol$Age.at.transfer[i] >= 35 & estrodiol$Age.at.transfer[i] < 40) {
    age_transfer_cat[i] = "35-39"
  } else if (estrodiol$Age.at.transfer[i] >= 40) {
    age_transfer_cat[i] = ">=40"
  }
}

head(age_transfer_cat)
length(age_transfer_cat)

estrodiol["age_transfer_cat"] = age_transfer_cat
```


## Basic Statistics

I will now run some basic statistics. 

```{r}
#Now let's run some basic stats
mean((estrodiol$Age.at.transfer))
sd((estrodiol$Age.at.transfer))


mean((estrodiol$Estradiol.value))
sd((estrodiol$Estradiol.value))
```

Now we will do some counting statistics. We are going to split our data into groups A and B. Group A has an estradiol level of less than 200.

```{r}
#Now we will do some counting statistics. We are going to split our data into groups A and B. Group A has an estradiol level of less than 200. 

row_a = which(estrodiol$Estradiol.value < 200)
group_a = estrodiol[row_a,]
nrow(group_a)
#There are 134 patients in group A

row_b = which(estrodiol$Estradiol.value >= 200)
group_b = estrodiol[row_b,]
nrow(group_b)
#There are 307 patients in group B
```

We will now examine the column called viable IUP. If that columns says yes, that means the patient is pregnant. We want to know the clinical prgnancy rate in total, and for groups A and B. 

```{r}
#We will now examine the column called viable IUP. If that columns says yes, that means the patient is pregnant. We want to know the clinical prgnancy rate

class(estrodiol$Viable.IUP.)
unique(estrodiol$Viable.IUP.)

yes_rows = which(estrodiol$Viable.IUP. == "Yes")
length(yes_rows)
#65% of the patients were pregnant

no_rows = which(estrodiol$Viable.IUP. == "No")
length(no_rows)
#35% of the patients were pregnant

#Now, we want to know what percent of patients from groups A and B had a positive preganacy. 
yes_rows_a = which(group_a$Viable.IUP. == "Yes")
length(yes_rows_a)
#58% of the patients in group A were pregnant

no_rows_a = which(group_a$Viable.IUP. == "No")
length(no_rows_a)
#42% of the patients in group A were pregnant

yes_rows_b = which(group_b$Viable.IUP. == "Yes")
length(yes_rows_b)
#68% of the patients in group B were pregnant

no_rows_b = which(group_b$Viable.IUP. == "No")
length(no_rows_b)
#32% of the patients in group B were pregnant
```

We will know row a chisq test for independence for groups A and B. I will first add groups A and B to the original dataframe
```{r}
group = array()
for (i in 1:nrow(estrodiol))
  if (estrodiol$Estradiol.value[i] < 200) {
    group[i] = "A"
  } else {
    group[i] = "B"
  }
  
length(group)
estrodiol["group"] = group
```

To check my work, I will create a contingency table
```{r, warning = FALSE}
library(MASS)       # load the MASS package 
tbl = table(estrodiol$group, estrodiol$Viable.IUP.) 
tbl                 # the contingency table
```

Now I will run the chisq test
```{r}
chisq.test(tbl) 
```


I will now perform the same operations with live birth columns (LB)
```{r}
#1 corresponds to a live birth. I will first see the total number of 1's
live_birth_rows = (which(estrodiol$LB == 1))
length(live_birth_rows)
#In total, there were 264 live births. 

still_birth_rows = (which(estrodiol$LB == 0))
length(still_birth_rows)
#In total, there were 170 non-live births.This does not add up to the total number of rows because some live birth data is missing. 

#Now I will create a contingency table for life births for groups A and B. 
tbl2 = table(estrodiol$group, estrodiol$LB) 
tbl2                 # the contingency table

chisq.test(tbl2) 
```
The findings for live birth for groups A and B are significantly different with an alpha of 0.05

## Demographic Statistics

Now we will dive in to some of the demographic statistics. 

```{r}
mean_age_a = mean(group_a$Age.at.transfer)
print(paste0("mean age for group A is: ", mean_age_a))

mean_age_b = mean(group_b$Age.at.transfer)
print(paste0("mean age for group B is: ", mean_age_b))


mean(estrodiol$BMI)
sd(estrodiol$BMI)
mean_bmi_a = mean(group_a$BMI)
print(paste0("mean BMI for group A is: ", mean_bmi_a))

mean_bmi_b = mean(group_b$BMI)
print(paste0("mean BMI for group B is: ", mean_bmi_b))



mean_end_tot = mean(estrodiol$Endometrial.thickness..mm.)
print(paste0("mean endometrial thickness in total is: ", mean_end_tot))

mean_end_a = mean(group_a$Endometrial.thickness..mm.)
print(paste0("mean endometrial thickness for group A is: ", mean_end_a))

mean_end_b = mean(group_b$Endometrial.thickness..mm.)
print(paste0("mean endometrial thickness for group B is: ", mean_end_b))
```

```{r}
unique(estrodiol$Smoker)

never_smoke_rows = which(estrodiol$Smoker == "Never")
tot_never_smoker = length(never_smoke_rows)
tot_never_smoker
former_smoke_rows = which(estrodiol$Smoker == "Former")
tot_former_smoker = length(former_smoke_rows)
tot_former_smoker

never_smoke_rows_a = which(group_a$Smoker == "Never")
num_never_smoker_a = length(never_smoke_rows_a)
num_never_smoker_a
former_smoke_rows_a = which(group_a$Smoker == "Former")
num_former_smoker_a = length(former_smoke_rows_a)
num_former_smoker_a

never_smoke_rows_b = which(group_b$Smoker == "Never")
num_never_smoker_b = length(never_smoke_rows_b)
num_never_smoker_b
former_smoke_rows_b = which(group_b$Smoker == "Former")
num_former_smoker_b = length(former_smoke_rows_b)
num_former_smoker_b

```

```{r}
unique(estrodiol$Nulliparous)

null1_rows = which(estrodiol$Nulliparous == 1)
tot_null1 = length(null1_rows)
tot_null1
null0_rows = which(estrodiol$Nulliparous == 0)
tot_null0 = length(null0_rows)
tot_null0

null1_rows_a = which(group_a$Nulliparous == 1)
tot_null1_a = length(null1_rows_a)
tot_null1_a
null0_rows_a = which(group_a$Nulliparous == 0)
tot_null0_a = length(null0_rows_a)
tot_null0_a

null1_rows_b = which(group_b$Nulliparous == 1)
tot_null1_b = length(null1_rows_b)
tot_null1_b
null0_rows_b = which(group_b$Nulliparous == 0)
tot_null0_b = length(null0_rows_b)
tot_null0_b

never_smoke_rows_b = which(group_b$Smoker == "Never")
num_never_smoker_b = length(never_smoke_rows_b)
num_never_smoker_b
former_smoke_rows_b = which(group_b$Smoker == "Former")
num_former_smoker_b = length(former_smoke_rows_b)
num_former_smoker_b

```

```{r}
unique(estrodiol$Race..choice.White.)
unique(estrodiol$Race..choice.East.Asian.)
unique(estrodiol$Race..choice.South.Asian.)

tot_white_row = which(estrodiol$Race..choice.White. == "Checked")
length(tot_white_row)
white_rowa = which(group_a$Race..choice.White. == "Checked")
length(white_rowa)
white_rowb = which(group_b$Race..choice.White. == "Checked")
length(white_rowb)

tot_nonwhite_row = which(estrodiol$Race..choice.White. == "Unchecked")
length(tot_nonwhite_row)
nonwhite_rowa = which(group_a$Race..choice.White. == "Unchecked")
length(nonwhite_rowa)
nonwhite_rowb = which(group_b$Race..choice.White. == "Unchecked")
length(nonwhite_rowb)


tot_e_asian_row = which(estrodiol$Race..choice.East.Asian. == "Checked")
length(tot_e_asian_row)
e_asian_rowa = which(group_a$Race..choice.East.Asian. == "Checked")
length(e_asian_rowa)
e_asian_rowb = which(group_b$Race..choice.East.Asian. == "Checked")
length(e_asian_rowb)

tot_none_asian_row = which(estrodiol$Race..choice.East.Asian. == "Unchecked")
length(tot_none_asian_row)
none_asian_rowa = which(group_a$Race..choice.East.Asian. == "Unchecked")
length(none_asian_rowa)
none_asian_rowb = which(group_b$Race..choice.East.Asian. == "Unchecked")
length(none_asian_rowb)


tot_s_asian_row = which(estrodiol$Race..choice.South.Asian. == "Checked")
length(tot_s_asian_row)
s_asian_rowa = which(group_a$Race..choice.South.Asian. == "Checked")
length(s_asian_rowa)
s_asian_rowb = which(group_b$Race..choice.South.Asian. == "Checked")
length(s_asian_rowb)

tot_nons_asian_row = which(estrodiol$Race..choice.South.Asian. == "Unchecked")
length(tot_nons_asian_row)
nons_asian_rowa = which(group_a$Race..choice.South.Asian. == "Unchecked")
length(nons_asian_rowa)
nons_asian_rowb = which(group_b$Race..choice.South.Asian. == "Unchecked")
length(nons_asian_rowb)
```

I am also going to add a column into the data set which will be used in the later analysis. Whites and all Asians will be categorized by 1. Non-whites/asians will be categorized by 0. 
```{r}
asian_white = array()

for (i in 1:nrow(estrodiol)) {
  if (estrodiol$Race..choice.East.Asian.[i] == "Checked") {
    asian_white[i] = 1
  } else if (estrodiol$Race..choice.South.Asian.[i] == "Checked") {
    asian_white[i] = 1
  } else if (estrodiol$Race..choice.White.[i] == "Checked") {
    asian_white[i] = 1
  } else {
    asian_white[i] = 0
  }
}

summary(asian_white)
sum(asian_white)

estrodiol["asian_white"] = asian_white
```

Now I will run t tests to compare these values for group A and B. I will begin with age. 
```{r}
#I will first test to see that these groups are normally distributed. 
qqnorm(group_a$Age.at.transfer)
qqline(group_a$Age.at.transfer)

qqnorm(group_b$Age.at.transfer)
qqline(group_b$Age.at.transfer)
#Both A and B look normal. 

t.test(group_a$Age.at.transfer, group_b$Age.at.transfer, alternative = "two.sided", var.equal = FALSE)
```

I will continue with BMI. 
```{r}
#I will first test to see that these groups are normally distributed. 
qqnorm(group_a$BMI)
qqline(group_a$BMI)
shapiro.test(group_a$BMI)

qqnorm(group_b$BMI)
qqline(group_b$BMI)
shapiro.test(group_b$BMI)
#These two groups are definitely not normally distributed. They are very much skewed right. 

t.test(group_a$BMI, group_b$BMI, alternative = "two.sided", var.equal = FALSE)

```
We see from the plots that group B is heavier than group A, which we will control in the logistic regression. 


I will continue with endometrial thickness 
```{r}
#I will first test to see that these groups are normally distributed. 
qqnorm(group_a$Endometrial.thickness..mm.)
qqline(group_a$Endometrial.thickness..mm.)
shapiro.test(group_a$Endometrial.thickness..mm.)

qqnorm(group_b$Endometrial.thickness..mm.)
qqline(group_b$Endometrial.thickness..mm.)
shapiro.test(group_b$Endometrial.thickness..mm.)
#These two groups are definitely not normally distributed. They are very much skewed right. 

t.test(group_a$Endometrial.thickness..mm., group_b$Endometrial.thickness..mm., alternative = "two.sided", var.equal = FALSE)

```
The p-value is large so we fail to reject the null hypothesis. 

## Logistic Regression

Let's now run some multiple logistic regression. 

```{r}
glm1 = glm(Viable.IUP. ~ BMI + Endometrial.thickness..mm. + 
            Age.at.transfer + Smoker + Estradiol.value
             , data = estrodiol, family = binomial)
summary(glm1)
```

```{r}
glm1 = glm(Viable.IUP. ~ group + BMI + Age.at.transfer, data = estrodiol, family = binomial)
summary(glm1)

exp(glm1$coefficients) #odds-ratios
exp(confint(glm1)) #confidence intervals around the odds-ratios
```
We see from above that our statistically significant variables are group and age at trasnfer

```{r}
glm2 = glm(Viable.IUP. ~ Age.at.transfer + group + 
             BMI + Smoker + Endometrial.thickness..mm. +
             Grade..1.good. + Nulliparous +
             asian_white, data = estrodiol, family = binomial)
summary(glm2)

exp(glm2$coefficients)
exp(confint(glm2))
```

After some back and forth, the researcher wished to study the response viable IUP with the following predictors. The only change from above is that we broke the age ranges into groups, so age is no longer a continuous variable. 
```{r}
glm3 = glm(Viable.IUP. ~ age_transfer_cat + group + 
             BMI + Smoker + Endometrial.thickness..mm. +
             Grade..1.good. + Nulliparous +
             asian_white, data = estrodiol, family = binomial)
summary(glm3)

exp(glm3$coefficients)
exp(confint(glm3))
```


Now we will repeat this analysis, but with live birth as the response. 
```{r}
glm4 = glm(LB ~ age_transfer_cat + group + 
             BMI + Smoker + Endometrial.thickness..mm. +
             Grade..1.good. + Nulliparous +
             asian_white, data = estrodiol, family = binomial)
summary(glm4)

exp(glm4$coefficients)
exp(confint(glm4))
```